Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-09-13 popl-rebuttal.md

Firstly, we thank the reviewers for their valuable feedback.

1. Read Uncommitted (RU), while a valid ANSI SQL isolation level, is
   often not implemented by the databases. For eg., Oracle supports
   only RC and SER, while Postgres maps RU to RC. Even when RU is
   implemented, it is never the default isolation level (Bailis et al,
   HotOS'13), and applications almost never demote the isolation level
   in practice (Bailis et al, SIGMOD'15). Thus, RC can be considered
   as a baseline isolation level that meets the performance demands of
   real-world applications. Lastly, RU violates atomicity, the basic
   tenet underlying transactions, and accommodating it would
   significantly complicate the formal framework. Weighing such pros
   and cons, we decided to elide RU.

2. While our language only includes the four basic SQL operations,
   they are nonetheless enough to implement database community's
   standard benchmarks, such TPC-C and TPC-E. Moreover, MySQL's
   (developer) documentation seems to indicate that its core calculus
   for SQL is indeed composed of these four basic operations, making
   our technique likely applicable on MySQL's core language, if not
   its surface language. We haven't analyzed other databases, but if
   the need arises, we can accommodate database-specific semantics as
   elaboration rules on the lines described in Sec. 6.1 para 1.

3. About our framework not admitting all SI executions, it seems that
   reviewer might have misunderstood our specification for SI. The
   spec allows concurrent transactions that do not conflict with the
   current SI transaction to commit concurrently, since SI's I_c spec
   (p.11) allows Δ to change as long as it doesn't conflict with δ.

4. We have ignored heap memory to focus on the database state in this
   paper. N Swamy et al (PLDI'13) have proposed Dijkstra monad to
   reason about heap memory monadically, and it would be interesting
   to see if Diskstra and DB monads can be cleanly composed to reason
   about the combination of heap and database.

### 2017-09-13 rebuttal.txt

We thank the reviewers for their valuable feedback.

Reviewer (1):
-------------

- No support for Read Uncommitted (RU) isolation.  While RU is a valid
  ANSI SQL isolation level, it is typically not implemented by modern
  databases. For example, Oracle supports only RC and SER, while
  Postgres maps RU to RC. Even when RU is supported, it is never the
  default isolation level (Bailis et al, HotOS'13)., and applications
  rarely demote isolation levels to RU in practice (Bailis et al,
  SIGMOD'15). Thus, RC can be considered as a baseline isolation level
  that meets the performance demands of real-world applications.
  Furthermore, RU violates atomicity, the basic tenet underlying
  transactions.  Given its lack of acceptance in the database
  community, and the complications it would introduce to our
  formalization, we opted to elide it from the development.

- Limited set of operations in the DSL.  While it is true our language
  only includes four basic SQL operations, they have proven sufficient
  to implement widely accewpted standard benchmarks, such TPC-C and
  TPC-E.  We also note that MySQL's developer documentation (see:
  ....) indicates that its core calculus for SQL is indeed
  composed of these four basic operations, making our technique likely
  applicable on MySQL's core language, if not its surface language. We
  expect similar conclusions to hold on other databases.

Reviewer (2):
-------------

- Not admitting valid snapshot isolation (SI) executions.  We believe
  the reviewer has misunderstood our specification for SI as $I_{ss}$,
  which only describes the semantics of "executing against a
  snapshot". SI indeed executes against a snapshot (thus its $I_e =
  I_{ss}$), but it *commits* against any non-write-write-conflicting
  database state (thus its $I_c = I_{ww}$). This is explained on p.13
  ("Snapshot Isolation").  The presented specification indeed allows
  non-conflicting concurrent transactions to commit in parallel.

- Decidability.  (a) The RG-TXN (Sec. 5.2, p. 20) rule does not lead
  to second-order quantification or inversion of quantifier positions.
  First, recall that each premise of the rule generates an independent
  verification condition (VC) dispatched to the solver. The VC
  corresponding to $\mathbf{F}\Rightarrow G$ first introduces $\delta$
  and $\Delta$ into the context as some sets (uninterpreted
  functions).  Next, it asserts the logical encoding of $F$ computed
  using the rules in Fig.10. Finally it asserts the negation of $G$,
  and checks for satisfiability. Note that in this process: (i). sets
  or functions are never quantified (thus no second-order
  quantification), and (ii). $\mathbf{F}$ is never negated (thus no
  inversion of quantifiers in $\mathbf{F}$). Other premises are
  similarly encoded.

  The reviewer is correct in observing that our encoding is not in the
  EPR fragment of FOL since we admit formulas of the form $\forall
  x.\exists y.\phi$.  However, our encoding is still decidable as it
  belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,
  that admits sentences of the form: $\exists^{}\forall^{2}\exists^{}$
  (see: Borger et al, The Classical Decision Problem, p.261; Gurevich,
  On the Classical Decision Problem, p.7); formulas generated by our
  encoding clearly belong to this fragment.  We will clarify this
  issue in the final version.

Reviewer (3);
-------------

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

### 2017-09-13 popl-rebuttal.md

Firstly, we thank the reviewers for their valuable feedback.

1. Read Uncommitted (RU), while a valid ANSI SQL isolation level, is
   often not implemented by the databases. For eg., Oracle supports
   only RC and SER, while Postgres maps RU to RC. Even when RU is
   implemented, it is never the default isolation level (Bailis et al,
   HotOS'13), and applications almost never demote the isolation level
   in practice (Bailis et al, SIGMOD'15). Thus, RC can be considered
   as a baseline isolation level that meets the performance demands of
   real-world applications. Lastly, RU violates atomicity, the basic
   tenet underlying transactions, and accommodating it would
   significantly complicate the formal framework, jeopardizing
   automation. Weighing such pros and cons, we decided to elide RU.

2. While our language only includes the four basic SQL operations,
   they are nonetheless enough to implement database community's
   standard benchmarks, such TPC-C and TPC-E. Moreover, MySQL's
   [developer documentation](https://dev.mysql.com/doc/internals/en/)
   seems to indicate that its core calculus for SQL is indeed composed
   of these four basic operations, making our technique likely
   applicable on MySQL's core language, if not its surface language.
   We haven't analyzed other databases, but if the need arises, we can
   accommodate database-specific semantics as elaboration rules on the
   lines described in Sec. 6.1 para 1.

3. About our framework not admitting all SI executions, the reviewer has 
   possibly misunderstood our specification for SI as
   $I_{ss}$, which only describes the semantics of "executing
   against a snapshot". SI indeed executes against a snapshot (thus its $I_e
   = I_{ss}$), but it commits against any non-write-write-conflicting
   database state (thus its $I_c = I_{ww}$). This is explained under
   "Snapshot Isolation" heading on p.13. Observe that the spec allows
   non-conflicting concurrent transactions to commit in parallel.

4. The RG-TXN rule does not lead to second-order quantification or
   inversion of quantifier positions for the reasons described below.
   Firstly, lets recall that each premise of the rule generates an
   independent verification condition (VC) dispatched to the solver.
   The VC corresponding to $F\Rightarrow G$ first introduces $\delta$
   and $\Delta$ into the context as *some* sets (uninterpreted
   functions, to be precise). Next, it asserts the logical encoding of
   $F$ computed using the rules in Fig.10. Finally it asserts the
   negation of $G$, and checks for satisfiability. Note that in this
   process: (a). Sets or functions are never quantified (thus no
   second-order quantification), and (b). $F$ is never negated (thus
   no inversion of quantifiers in $F$). Other premises are similarly
   encoded.

5. We have been a bit sloppy about describing the fragment of FOL that
   our logic belongs to. The reviewer is indeed right that $\forall
   x.\exists y.\phi$ doesn't belong to EPR. The appropriate fragment
   in our case is a closely related fragment called
   Godel-Kalmar-Schutte, which admits sentences of form
   $\exists^{*}\forall^{2}\exists^{*}$, while remaining decidable
   (Borger et al, The Classical Decision Problem, p.261; Gurevich, On
   the Classical Decision Problem, p.7). We apologize for not being
   precise, and will promptly correct the mistake.

6. We have ignored heap memory to focus on the database state in this
   paper. N Swamy et al (PLDI'13) have proposed Dijkstra monad to
   reason about heap memory monadically, and it would be interesting
   to see if Dijkstra and DB monads can be cleanly composed to reason
   about the combination of heap and database.

### 2017-09-13 evaluation.tex

\subsection{Evaluation}
\label{sec:case-studies}

In this section, we present our experience in running \tool on two different applications: \emph{Courseware}: a
course registration system described by~\cite{gotsmanpopl16}, and
% 2. RUBiS: an eBay-like auction site~\cite{rubis} that allows users to
% bid for items on sale, and pay for items from a wallet modeled after a
% bank account, and 
the \emph{TPC-C} benchmark.

@@

\begin{table}[]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
                                 & \multicolumn{1}{l|}{\C{new\_order}} & \multicolumn{1}{l|}{\C{delivery}} & \multicolumn{1}{l|}{\C{payment}} & \multicolumn{1}{l|}{\C{order\_status}} & \multicolumn{1}{l|}{\C{stock\_level}} \\ \hline
\multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
\multicolumn{1}{|l|}{PostgreSQL} & RR                                    & RR                                  & RC                                 & RC                                       & RC                                      \\ \hline
\end{tabular}
\caption{The discovered isolation levels for TPC-C transactions}
\label{tab:tpcc}
\end{table}

@@

To sanity-check the results of \tool, we conducted experiments with a
high-contention OLTP workload on TPC-C aiming to explore the space of
correct isolation levels for different transactions. The workload
involves a mix of all five TPC-C transactions executing against a
TPC-C database with 10 warehouses. Each warehouse has 10 districts,
and each district serves 3000 customers. There are a total of 5
transactions in TPC-C, and given that MySQL and PostgreSQL support 3
isolation levels each, there are a total of $3^5 = 243$ different
configurations of isolation levels for TPC-C transactions on MySQL and
PostgreSQL. We executed the benchmark with all 243 configurations, and
found 171 of them violated at least one of the four invariants we
considered.  Table~\ref{fig:tpcc-eval} shows the summary of these
results. As expected, the isolation levels that \tool infers for the
TPC-C transactions do not result in invariant violations, either on
MySQL or on PostgreSQL, and were determined to be the weakest safe
assignments possible.

>>>>

\section{Evaluation}
\label{sec:case-studies}

In this section, we present our experience in running \tool on two
different applications: \emph{Courseware}: a course registration
system described by~\cite{gotsmanpopl16}, and the \emph{TPC-C}
benchmark.

@@

\begin{table}[]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
                                 & \multicolumn{1}{l|}{\C{new\_order}} & \multicolumn{1}{l|}{\C{delivery}} & \multicolumn{1}{l|}{\C{payment}} & \multicolumn{1}{l|}{\C{order\_status}} & \multicolumn{1}{l|}{\C{stock\_level}} \\ \hline
\multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
\multicolumn{1}{|l|}{PostgreSQL} & SI                                    & SI                                  & RC                                 & RC                                       & RC                                      \\ \hline
\end{tabular}
\caption{The discovered isolation levels for TPC-C transactions}
\label{tab:tpcc}
\end{table}

@@

To sanity-check the results of \tool, we conducted experiments with a
high-contention OLTP workload on TPC-C aiming to explore the space of
correct isolation levels for different transactions. The workload
involves a mix of all five TPC-C transactions executing against a
TPC-C database with 10 warehouses. Each warehouse has 10 districts,
and each district serves 3000 customers. There are a total of 5
transactions in TPC-C, and given that MySQL and PostgreSQL support 3
isolation levels each, there are a total of $3^5 = 243$ different
configurations of isolation levels for TPC-C transactions on MySQL and
PostgreSQL. We executed the benchmark with all 243 configurations, and
found 171 of them violated at least one of the four invariants we
considered.  As expected, the isolation levels that \tool infers for the
TPC-C transactions do not result in invariant violations, either on
MySQL or on PostgreSQL, and were determined to be the weakest safe
assignments possible.

### 2017-09-13 implementation.tex

type district = {d_id: int; d_next_o_id: int}
type order = {o_id: int; o_d_id: int; o_c_id: int; o_ol_cnt: int}
type order_line = {ol_o_id: int; ol_d_id: int; ol_i_id: int; ol_qty: int}
type stock = {s_i_id: int; s_d_id:int; s_qty: int}
\end{ocaml}
\caption{OCaml type definitions corresponding to the TPC-C schema from
Fig.~\ref{fig:schema1}}
\label{fig:ocaml-schema}
\end{figure}

We have implemented our DSL to define transactions as monadic
computations in OCaml (modulo some syntactic sugar), and our automatic
reasoning framework as an extra frontend pass (called \tool) in the ocamlc
4.03 compiler\footnote{The source code is available at available at
\emph{\ldots}}. The input to \tool is a program in our DSL that
describes the schema of the database as a collection of OCaml type
definitions, and transactions as OCaml functions, whose top-level
expression is an application of the \C{atomically\_do} combinator to a
function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema} can
be described via the OCaml type definitions shown in
Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
the program in the form of a collection of guarantees ($G$), one per
transaction, and an invariant $I$ that is a conjunction of the
integrity constraints on the database. An auxiliary DSL that includes
the first-order logic (FOL) combinators has been implemented for this
purpose. \tool's verification pass follows OCaml's type checking pass,
hence the concrete artifact of verification is OCaml's typed AST. The
tool is already equipped with  an axiomatization of PostgreSQL and
MySQL's isolation levels expressed in our FOL DSL. Other data stores
can be similarly axiomatized. The concrete result of verification is
an assignment of an isolation level of the selected data store to each
transaction in the program.

@@

\tool uses the Z3 SMT solver as its underlying reasoning engine. Each
implication check described above is first encoded in FOL, applying
the translation described in Sec.~\ref{sec:inference} wherever
necessary.

@@

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in Sec.~\ref{sec:isolation} leave out
certain nuances of their implementations on real data stores, which
need to be taken into account for verification to be effective in
practice.  We consider these into account while linking \tool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgreSQL implements an \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is SI, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if they still exist), write locks are obtained, and
the update is performed, provided that the record still meets
\C{UPDATE}'s search criteria. If the record no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked until the transaction commits.

>>>>

type district = {d_id: int; d_next_o_id: int}
type order = {o_id: int; o_d_id: int; o_c_id: int; o_ol_cnt: int}
type order_line = {ol_o_id: int; ol_d_id: int; ol_i_id: int; ol_qty: int}
type stock = {s_i_id: int; s_d_id:int; s_qty: int}
\end{ocaml}
\caption{OCaml type definitions corresponding to the TPC-C schema from
Fig.~\ref{fig:schema}}
\label{fig:ocaml-schema}
\end{figure}

We have implemented our DSL to define transactions as monadic
computations in OCaml (modulo some syntactic sugar), and our automatic
reasoning framework as an extra frontend pass (called \tool) in the ocamlc
4.03 compiler\footnote{The source code is available at available at
\emph{https://goo.gl/nmUXZK}}. The input to \tool is a program in our DSL that
describes the schema of the database as a collection of OCaml type
definitions, and transactions as OCaml functions, whose top-level
expression is an application of the \C{atomically\_do} combinator to a
function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema} can
be described via the OCaml type definitions shown in
Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
the program in the form of a collection of guarantees ($G$), one per
transaction, and an invariant $I$ that is a conjunction of the
integrity constraints on the database. An auxiliary DSL that includes
the first-order logic (FOL) combinators has been implemented for this
purpose. \tool's verification pass follows OCaml's type checking pass,
hence the concrete artifact of verification is OCaml's typed AST. The
tool is already equipped with  an axiomatization of PostgreSQL and
MySQL's isolation levels expressed in our FOL DSL. Other data stores
can be similarly axiomatized. The concrete result of verification is
an assignment of an isolation level of the selected data store to each
transaction in the program.

@@

\tool uses the Z3 SMT solver as its underlying reasoning engine. Each
implication check described above is first encoded in FOL, applying
the translation described in \S\ref{sec:inference} wherever
necessary.

@@

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in \S\ref{sec:isolation} leave out
certain nuances of their implementations on real data stores, which
need to be taken into account for verification to be effective in
practice.  We consider these into account while linking \tool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgreSQL implements an \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is SI, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if they still exist), write locks are obtained, and
the update is performed, provided that the record still meets
\C{UPDATE}'s search criteria. If the record no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked until the transaction commits.

### 2017-09-13 introduction.tex

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules would likely discourage applications
programmers from using them, we present an inference procedure based
on a state transformer semantics that automatically verifies the
weakest isolation level for a transaction that ensures its consistency
requirements are maintained.  The key to inference is a novel formulation
of database state (represented as sets of tuples) as a monad, and in
which database computations are interpreted as transformers over these
sets.  This interpretation leads to an encoding of database
computations amenable for verification by off-the-shelf SMT solvers.
The paper makes the following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
\item We analyze properties of weak isolation in the context of an
  embedded DSL for OCaml that treats SQL-based relational database
  operations (e.g., inserts, selects, deletes, joins, updates, etc.)
  as monadic computations over an abstract database state.
 \item We develop an operational semantics and compositional proof
   system for this language capable of relating high-level application
   invariants to database state, parameterized by a weak isolation
   semantics that selectively exposes the visibility of these
   operations to other transactions.
 \item Exploiting the monadic definition of database state, we devise
   an inference algorithm capable of discovering the weakest isolation
   level that is sound with respect to a transaction's high-level
   consistency requirements.  The algorithm interprets database
   operations as state transformers expressed in a language amenable
   for translation into a decidable fragment of first-order logic, and
   is thus suitable for automated verification using off-the-shelf SMT
   solvers.
 \item We present details of a realistic implementation along with an
   evaluation study on real-world database benchmarks that justify our
   approach and the utility of our proof methodology and inference
   mechanism..
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

The remainder of the paper is organized as follows. The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:inference} describes the inference
algorithm, and the state transformer encoding.  We describe our
implementation, benchmark results, and case studies in
\S\ref{sec:case-studies}.  Related work and conclusions are given in
\S\ref{sec:relatedwork}.

>>>>

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules would likely discourage applications
programmers from using them, we present an inference procedure based
that automatically verifies the weakest isolation level for a
transaction that ensures its consistency requirements are maintained.
The key to inference is a novel formulation of database state
(represented as sets of tuples) as a monad, and in which database
computations are interpreted as state transformers over these sets.
This interpretation leads to an encoding of database computations
amenable for verification by off-the-shelf SMT solvers.  The paper
makes the following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
\item We analyze properties of weak isolation in the context of an
  embedded DSL for OCaml that treats SQL-based relational database
  operations (e.g., inserts, selects, deletes, joins, updates, etc.)
  as monadic computations over an abstract database state.
 \item We develop an operational semantics and compositional proof
   system for this language capable of relating high-level application
   invariants to database state, parameterized by a weak isolation
   semantics that selectively exposes the visibility of these
   operations to other transactions.
 \item Exploiting the monadic definition of database state, we devise
   an inference algorithm capable of discovering the weakest isolation
   level that is sound with respect to a transaction's high-level
   consistency requirements.  The algorithm interprets database
   operations as state transformers expressed in a language amenable
   for translation into a decidable fragment of first-order logic, and
   is thus suitable for automated verification using off-the-shelf SMT
   solvers.
 \item We present details of a realistic implementation along with an
   evaluation study on real-world database benchmarks that justify our
   approach and the utility of our proof methodology and inference
   mechanism..
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

The remainder of the paper is organized as follows. The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:inference} describes the inference
algorithm, and the state transformer encoding.  We describe our
implementation in \S\ref{sec:implementation}, and provide case studies
and benchmark results in \S\ref{sec:case-studies}.  Related work is
given in \S\ref{sec:relatedwork}, and \S\ref{sec:conclusions} presents
conclusions.
